Numerical investigation of two-phase flow in high-permeability porous media Effect of permeability variation on traction terms

### 2020-03-30 draft.tex
minor

Due to the larger pore size, two-phase flow in high-permeability porous
media (hereafter, high-permeability porous media refers to media for
which the characteristic particle diameter is about one millimeter and
above) results in a complex interaction between capillary, gravity and
viscous forces {[}\protect\hyperlink{ref-davit2018one}{4}{]}. Because of
this complex interplay, the characteristic flow regimes are very
different from those observed for surface tension dominated flow
{[}\protect\hyperlink{ref-dullien2012porous}{5}{]}. For low-permeability
medium, if the surface tension between the fluids is not too low and the
viscosity not too high, that is the capillary number is low (usually
inferior to \(10^{-3}\)), the fluid repartition pattern is described as
two independent flow streams separate by a multitude of stable meniscus
at steady state, as illustrated in \cref{fig:fluidPatterns} (a). In this
case, for a completely wet medium, the wetting phase flows in the small
pores while the non-wetting phase tends to occupy the larger pores. For
high permeability porous media, the fluid velocity can be high enough
that the viscous forces dominate over capillary forces (alike gravity
and inertial effects may become important if Bond and Reynolds numbers
are high, respectively). Then, the fluids repartition patterns can take
two forms, either the non-wetting phase is continuous, see for example
\cref{fig:fluidPatterns} (b), or is flowing as droplets or ganglia as in
\cref{fig:fluidPatterns} (c), while the wetting phase is flowing as a
film in contact with the solid. In both cases, the two fluids occupy
most of the pores at the same time and the non-wetting phase flows at
the center of the pores surrounded by the wetting phase. One can note
that these flow regimes are equivalent to those observed in two-phase
flows in a tube (slug flow, bubbly flow). Strictly speaking, these
different regimes must be considered when attempting to describe
two-phase flows with continuous macroscopic equations. Indeed, it has
been shown on numerous works that the overall flow depends on the flow
regimes
{[}\protect\hyperlink{ref-Avraam1995a}{6},\protect\hyperlink{ref-armstrong2016beyond}{7}{]}.
In the first glance, one would consider that the exchange terms between
the fluid phases as negligible compared to their counterpart between the
fluid phases and the solid phase for surface-tension dominated flwo and
that the extent of the fluid-fluid interface is small, as in
\cref{fig:fluidPatterns} (a). On the other hand, and this is what we are
interested in here, this is not necessarily the case for regimes
specific to flow in high-permeability porous media for which the extent
of the fluid-fluid interface is large. This is important because, as we
will see in the next section, these exchange terms between phases are
the basis of any attempt to establish continuous relationships on a
macroscopic scale starting from the pore scale.

@@

Here we present two different approaches to treat the continuous
macroscopic momentum equations for two-phase flow. Although
fundamentally identical, these approaches were developed for the needs
of two different specialties, namely, the study of groundwater flows and
the study of the hydrodynamic of chemical reactors.

\hypertarget{porous-media-approach}{%
\subsubsection{Porous media approach}\label{porous-media-approach}}

The ubiquitous continuous model used to describe two-phase flows in
soils is based on an extension of the Darcy equation (creeping saturated
single-phase flow in porous media). The whole model, also known as
Muskat equations
{[}\protect\hyperlink{ref-wyckoff1936flow}{8},\protect\hyperlink{ref-muskat1938flow}{9}{]},
reads

@@

In these equations, the subscripts \(o\) and \(w\) refer to the
non-wetting and wetting phase, respectively. \(S_{i}\) is the saturation
of phase \(i\), \(\varepsilon\) is the medium porosity,
\(\mathbf{U}_{i}\) is the superficial mean velocity of fluid \(i\),
\(P_i\) is the intrinsic mean pressure of fluid \(i\) and \(\mathbf{K}\)
is the absolute permeability tensor. The generalization toward two-phase
flows involves the introduction of the relative permeability terms
\(k_{ri}\) which account for one phase for the inaccessible void space
occupied by the other fluid
{[}\protect\hyperlink{ref-dullien2012porous}{5}{]}. To close the set of
the macroscopic equations, the relative permeabilities are assumed to be
non-linear functions of the saturation and a relation between the
macroscopic pressure of each fluid has to be furnished. This relation,
is known as the capillary pressure relation and, as for the relative
permeabilities, is supposed to vary non-linearly with the saturation
only {[}\protect\hyperlink{ref-leverett1941capillary}{10}{]}.

A significant amount of work attempted to make improvements to the
Muskat equations (e.g.~including moving contact line
{[}\protect\hyperlink{ref-kalaydjian1992dynamic}{11}--\protect\hyperlink{ref-barenblatt2003mathematical}{13}{]}
or take into account the trapped phases
{[}\protect\hyperlink{ref-hilfer1998macroscopic}{14}{]}). Here we would
stress that the concept of relative permeabilities into the Muskat
equations is dedicated to independent flow pathways only
{[}\protect\hyperlink{ref-blunt2017multiphase}{15}{]} and the underlying
assumption that the two streams do not interfere with each other.
However, in case that the area of the fluid-fluid interface is not
negligible compared to the solid-fluid interface, as for flow regimes
\cref{fig:fluidPatterns} (b) and \cref{fig:fluidPatterns} (c), one would
include supplementary terms dedicated to the momentum exchange across
the fluid-fluid interface due to normal and shear stresses exerted on it
(i.e.~coupling between the phases) into the macroscopic momentum
equations. In fact, it has been theoretically established that the
macroscopic equations of motion obtained by averaging of the Stokes
equations contain momentum exchange terms between fluids
{[}\protect\hyperlink{ref-MARLE1982643}{16}--\protect\hyperlink{ref-kalaydjian1990origin}{19}{]}.
The volume averaged momentum equation for the \(i\)-phase read

The next step in the upscaling methods literature is to express the
boundaries integral of exchange between the phases in terms of averaged
quantity through the resolution of boundary value problems on a
representative periodic unit-cell. We need to go no further here but we
insist that, as shown in {[}\protect\hyperlink{ref-Lasseux1996}{20}{]},
once the closure problems have been solved, the macroscopic equations of
motion can be put into a form analogous to the generalized Darcy's law,

in which \(\mathbf{K}_{ij}^{*}\) are the coupled relative permeability
tensors.

Many studies have attempted to calculate coupled permeabilities
experimentally based on the +\textbf{???}, but before we turn to present
results from the literature, we discuss the chemical engineering point
of view on the continuous model.

\hypertarget{chemical-engineering-approach}{%
\subsubsection{Chemical engineering
approach}\label{chemical-engineering-approach}}

CFD modeling applied to the hydrodynamics of trickle-bed reactors (TBR)
is in most cases focused on an Euler-Euler flow formulation. The two
fluids are thus treated as two interpenetrating continua. The volume
averaged momentum equation for the two fluids to solve, taking into
account that the inertial effects are not negligible here, read

\[
\rho_{i}\varepsilon_{i}\left(\frac{\partial\mathbf{U}_{i}}{\partial t}+\mathbf{U}_{i}\cdot\nabla\mathbf{U}_{i}\right)=-\varepsilon_{i}\nabla P_{i}+\varepsilon_{i}\rho_{i}\mathbf{g}+\mathbf{F}_{mi}+\varepsilon_{i}\nabla\cdot\boldsymbol{\tau}_{i}+\nabla\cdot\mathbf{R}_{i},
\] \{\#eq:trickleMomentum\}

where \(\varepsilon_{i}\) is the volume fraction of phase \(i\), the
velocity and the pressure are averaged quantities identical to the
generalized Darcy equations, \(\boldsymbol{\tau}_{i}\) denotes the
viscous stress tensor and \(\mathbf{R}_{i}\) is the pseudo-turbulence
stress. The vectors \(\mathbf{F}_{mi}\) are the momentum transfer terms
at the interface for the \(i\)-phase. These terms have to be expressed
in terms of constitutive relation to close the set of equations, and the
main difference between the existing continuous models is the choice of
correlations used for these terms. We stress here that the two equations
+\textbf{???} and +\textbf{???} are equivalents (except for the
non-linear inertial terms).

\hypertarget{modeling-of-the-momentum-exchange-between-fluids}{%
\subsubsection{Modeling of the momentum exchange between
fluids}\label{modeling-of-the-momentum-exchange-between-fluids}}

Here we present different estimation of the momentum transfer terms
(i.e.~coupled relative permeabilities in the porous media framework)
between fluids. Based on the generalized Darcy law with cross terms
+\textbf{???}, experimental works computed the coupled transport
coefficient through steady-state cocurent flow in sandpack with
alternatively one fluid, then the other, which is submitted to a null
pressure gradient. This protocol was used with oil and water in a
cylindric sandpack
{[}\protect\hyperlink{ref-zarcone1994determination}{21}{]} and the
authors found a negligible effect of the coupled permeabilities in the
overall flow. With the same protocol, with oil and water in a
2D-sandpack Dullien and Dong
{[}\protect\hyperlink{ref-Dullien1996}{22}{]} found that the coupled
permeabilities are important since they can contribute at best to 35\%
of the effective permeability. Alternatively, authors imposed
{[}\protect\hyperlink{ref-ramakrishnan2015measurement}{23}{]} a null
displacement and measured the induced pressure drop, with air and water
in a Berea sandstone core they found that the coupled transport
coefficients must not be overlooked for intermediate saturations.
Several authors conducted experiments with two different set-ups as
proposed by Rose {[}\protect\hyperlink{ref-Rose1988}{24}{]}, e.g.
{[}\protect\hyperlink{ref-bentsen1993use}{25}{]}, in which the authors
made cocurrent and countercurrent experiments with water and oil in a
sandpack and found that coupled permeabilities reach, at least, 15\% of
the effective permeability value. However, it was pointed out that the
saturation between the two sets of experiments can be very different and
therefore the computed relative permeabilities can not be safely
compared {[}\protect\hyperlink{ref-langaas2001numerical}{26}{]}. The
effect of the non-wetting phase connectivity on the transport parameters
was extensively studied in {[}\protect\hyperlink{ref-Avraam1995}{27}{]},
the authors performed steady-state cocurrent two-phase flow in 2D-micro
model experiments and found that the contribution of the coupled
permeabilities on the flow is non-negligible and depend on the flow
regimes.

Several numerical studies investigate the share of the coupled
permeabilities in the effective permeability. The seminal work of
Rothman {[}\protect\hyperlink{ref-Rothman1990}{28}{]} examined the
question by conducting two-phase flow simulations in simple geometries
with the immiscible lattice-gas method. The author found non-negligible
participation of the coupled permeabilities by applying the volume force
alternatively on each fluid. Numerous authors used the lattice-Boltzmann
method, such as {[}\protect\hyperlink{ref-Li2005}{29}{]}, in which the
authors examined the value of the coupled permeabilities as a function
of the saturation in a 3D-sphere pack and they found results in
agreement with Rothman's results. In
{[}\protect\hyperlink{ref-Yiotis2007}{30}{]} the authors used a lattice
Boltzmann method in 2D and 3D pore networks and found a non-wetting
apparent relative permeability greater than unity when the wetting fluid
is more viscous than the non-wetting fluid. This result is a
manifestation of the lubrification effect, also observed in experiments
{[}\protect\hyperlink{ref-odeh1959effect}{31}{]}, and arising because of
a strongly hydraulic coupling between the fluids. Recently, in
{[}\protect\hyperlink{ref-shams2018study}{32}{]}, the authors have used
a Volume Of Fluid method to study the transport coefficients of fluid
layers in non-circular capillary tubes. Based on analogy with a model
Couette flow the authors have derived simple relations that can predict
with good accuracy the transport coefficients of fluid layers.

@@

In this study, we investigate the effect of the varying intrinsic
permeability and capillary number on the momentum transfer terms through
direct numerical simulations of two-phase flow in a model
high-permeability 2D-porous media. To do so we compute the surface
integrals in +\textbf{???} on each boundaries of the system. The idea of
solving a depth-averaged flow in a Hele-Shaw allows us to introduce an
additional permeability term and thus to vary the intrinsic permeability
of the strucure without having to modify the geometry per se.

@@

In the following, the superficial mean of a physical quantity
\(\theta_i\) associated with the \(i\)-phase is given by

@@

Macroscopic quantities such as saturation and intrinsic mean pressure
for two fluids \(i=w,o\) read

@@

As previously said, an essential step in order to derive some
macroscopic law at macroscale from the microscale is to use averaged
quantities to express the momentum transfer terms between phases that
arising because of the averaging theorem application
{[}\protect\hyperlink{ref-Whitaker2013}{33}{]}. Remember that these
closure relations are not presented here but rather we leave intact the
surface integrals, the averaged momentum balance equations +\textbf{???}
for two fluids \(w\) and \(o\), are recast as

@@

where \(\varepsilon_{i}\) and \(P_{i}\) are the macroscopic quantities
defined above, \(\mathbf{T}_{ij}\) the traction (name given to surface
forces in the Boundary Integral Methods literature in which these
integrals are central) exerted by the \(i\)-fluid on the \(j\)-fluid and
\(\mathbf{T}_{is}\) the traction exerted by the \(i\)-fluid on the solid
surfaces {[}\protect\hyperlink{ref-Kalaydjian1987}{34}{]}. The traction
terms are written as

@@

where \(\mathbf{n}_{ij}\) is the normal vector at the interface and
pointing toward the j-phase, \(p_{i}\) the pointwise pressure of the
i-fluid and
\(\mathbf{e}_{i}=\frac{1}{2}\left(\nabla\mathbf{u}_{i}+\left(\nabla\mathbf{u}_{i}\right)^{T}\right)\)
the rate-of-strain tensor for Newtonian fluids with \(\mathbf{u}_{i}\)
the pointwise velocity of the i-fluid.

@@

Two-phase flow between two parallel plates under study here is depicted
in \cref{fig:modelStudy}. This illustration represents a cocurrent flow
of two fluids inside a Hele-Shaw cell parallel to the \(x-y\) plane,
with the \(z\)-axis perpendicular to the plates, and for which \(h\) is
the aperture between the plates. A solid obstacle of circular
cross-section is sandwiched between the two plates. Tangential (cell
plane) stress, as well as stress arising from the perpendicular
confinement, need to be taking into account to properly describe the
flow. The important parameter here are the aspect ratio \(h/l\) where
\(l\) is some characteristic transverse length (hereafter \(l\) is the
cell width as shown in \cref{fig:modelStudy}) and the capillary number.

@@

where \(\gamma\) denotes the surface tension and \(\kappa_{\parallel}\)
the curvature in the Hele-Shaw plane
{[}\protect\hyperlink{ref-saffman1958penetration}{35}{]}. The
+\textbf{???} was completed in
{[}\protect\hyperlink{ref-park1984two}{36}{]} by an additional term that
pertains for dynamic film formation and a corrective \(\pi/4\) factor
for the curvature in the cell plane.

To compute the traction exerted by the fluids on the cell plates we need
to consider that the velocity profile in the \(z\)-direction is well
described by the Poiseuille flow profile,

@@

We can show that the average velocity taken over the interval between
the plates is written as follows

@@

There is therefore a direct analogy between flows in a 2D porous medium
and a flow in a Hele-Shaw cell, as can be seen in +\textbf{???}, where
the permeability is given by \(h^{2}/12\)
{[}\protect\hyperlink{ref-saffman1958penetration}{35}{]}.

@@

In this equation, \(\mathbf{u}\) is the velocity field, \(\tau\) and
\(\psi\) are two numerical parameters that control the diffuse interface
thickness and the amount of reinitialization of the \(\phi\) function,
respectively
{[}\protect\hyperlink{ref-Olsson2005}{37},\protect\hyperlink{ref-Olsson2007}{38}{]}.

\hypertarget{flow-model}{%
\subsubsection{Flow model}\label{flow-model}}

@@

where \(k=\frac{\sqrt{12}}{h}\) is an additional term that pertain to a
Darcy-term arising from the depth-averaging of the Stokes equation, as
described in the theoretical background section and \(\mathbf{F}_{st}\)
is the capillary contribution. The capillary contribution is difficult
to compute in volumetric methods such as Level Set and is itself the
object of a large body of work
{[}\protect\hyperlink{ref-Popinet2018}{39}{]}, this term is defined here
as

@@

where \(\gamma\) is the surface tension, \(\kappa\) is the interface
mean curvature, \(\mathbf{n}\) denotes the unit normal to the interface
and \(\delta(\phi)\) is the Dirac delta function localized on the
interface {[}\protect\hyperlink{ref-Galusinski2008}{40}{]}. In this
framework the unit normal vector as well as the mean curvature are
obtained with the level set function \(\phi\) as

@@

respectively, where the \(\pi/4\) correction was derived in
{[}\protect\hyperlink{ref-park1984two}{36}{]} and
\(\kappa_{\perp}=\frac{2}{h}\) is the out-of-plane curvature as
described previously in the theoretical background section.

@@

\hypertarget{validation}{%
\subsubsection{Validation}\label{validation}}

@@

\begin{longtable}[]{@{}lll@{}}
\caption{Equivalence between the value of the cell plates aperture and
the intrinsic permeability of the model obtained either from the
intrinsic permeability for flow between two plates or from single-phase
simulation in our model. \label{tbl:permeability}}\tabularnewline
\toprule
\(l/h\) & \(h^2/12\) (darcy) & \(k\) Simulation (darcy)\tabularnewline
\midrule
\endfirsthead
\toprule
\(l/h\) & \(h^2/12\) (darcy) & \(k\) Simulation (darcy)\tabularnewline
\midrule
\endhead
0.2 & \(5.21 \times 10^{5}\) & \(1.48 \times 10^{4}\)\tabularnewline
2.0 & \(5.21 \times 10^{3}\) & \(2.85 \times 10^{3}\)\tabularnewline
4.0 & \(1.30 \times 10^{3}\) & \(8.63 \times 10^{2}\)\tabularnewline
8.0 & \(3.26 \times 10^{2}\) & \(2.35 \times 10^{2}\)\tabularnewline
20 & \(5.21 \times 10^{1}\) & \(3.94 \times 10^{1}\)\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{figure}
\hypertarget{fig:drag_plates}{%
\centering
\includegraphics{figures/pdf/drag_plates.pdf}
\caption{\(x\)-component of the traction on Hele-Shaw cell plates by
each fluid in UC3, as a function of the dimensionless gap and for
different capillary numbers at steady-state.}\label{fig:drag_plates}
}
\end{figure}

>>>>

Due to the larger pore size, two-phase flow in high-permeability porous
media (hereafter, high-permeability porous media refers to media for
which the characteristic particle diameter is about one millimeter and
above) results in a complex interaction between capillary, gravity and
viscous forces {[}\protect\hyperlink{ref-davit2018one}{4}{]}. For
low-permeability porous media, the flow is usually dominated by surface
tension force, i.e.~the capillary number is low (usually inferior to
\(10^{-3}\)), and the fluid repartition pattern is well described as two
independent flow streams separate by a multitude of stable meniscus at
steady state, as illustrated in \cref{fig:fluidPatterns} (a)
{[}\protect\hyperlink{ref-dullien2012porous}{5}{]}. This regime is
characterized by the small extent of the fluid-fluid interface, as the
fluid phases are segregated, with the wetting phase flowing into the
larger pores while the wetting phase occupies the smaller pores. In
contrast, for high permeability porous media, the flow is no longer
capillary-dominated and the viscous forces become significant (also
gravity and inertial effects may become important if Bond and Reynolds
numbers are high, respectively). Thus the fluids repartition patterns
can take two forms, either the non-wetting phase is continuous, see as
an example \cref{fig:fluidPatterns} (b), or is flowing as droplets or
ganglia, as in \cref{fig:fluidPatterns} (c). For these two options, the
wetting phase is flowing as a film in contact with the solid and the
non-wetting phase flows at the center of the pores surrounded by the
wetting phase. One can note that these flow regimes are analogous to
those observed in gas-liquid flows in a tube (annular flow, bubbly flow,
etc.) {[}\protect\hyperlink{ref-taitel1976model}{6}{]}. Strictly
speaking, these different regimes must be considered when attempting to
describe two-phase flows with continuous macroscopic equations. Indeed,
it has been shown several times that the overall flow properties depend
on the flow regimes
{[}\protect\hyperlink{ref-Avraam1995a}{7},\protect\hyperlink{ref-armstrong2016beyond}{8}{]}.
At first glance, one would consider that the exchange terms between the
fluids, through their common surfaces, as negligible compared to their
counterpart between the fluid phases and the solid phase for
surface-tension dominated flow since the extent of the fluid-fluid
interface is small. On the other hand, and this is what we are
interested in here, this is not necessarily the case for regimes
specific to flow in high-permeability porous media for which the extent
of the fluid-fluid interface is large. This is important because, as
discussed in the next section, these exchange terms between phases are
the basis of any attempt to establish continuous relationships on a
macroscopic scale starting from the pore scale.

@@

Volume averaged microscopic equation of motion for two Newtonian fluids
\(i,j=o,w\:\mathrm{and}\:i\neq j\) reads

\[
\rho_{i}\varepsilon_{i}\left(\frac{\partial\mathbf{U}_{i}}{\partial t}+\mathbf{U}_{i}\cdot\nabla\mathbf{U}_{i}\right)=-\varepsilon_{i}\nabla P_{i}+\varepsilon_{i}\rho_{i}\mathbf{g}+\varepsilon_{i}\nabla\cdot\boldsymbol{\tau}_{i}+(\mathbf{F}_{ij}+\mathbf{F}_{is})
\] \{\#eq:averagedMomentum\}

where \(\varepsilon_{i}=V_i/V\) is the volume fraction of the
\(i\)-fluid and \(V\) is a representative elementary volume, \(U_i\) is
the intrinsic average velocity of the \(i\)-fluid, \(P_i\) is the
average pressure of the \(i\)-fluid, \(\boldsymbol{\tau}_{i}\) denotes
the average viscous stress tensor. The last two terms in the left hand
side denote the force of interaction per unite volume exerted by fluid
\(i\) upon fluid \(j\) and the force of interaction exerted by fluid
\(i\) upon the solid phase, respectively
{[}\protect\hyperlink{ref-Kalaydjian1987}{9}{]}.

From there, there are two possible ways to solve +\textbf{???} and the
associated equations. In volume averaging litterature, one seeks to
express the terms of interaction between the phases with averaged
quantities and to recast equation +\textbf{???} in a form similar to
Darcy's law (equation for creeping saturated single-phase flow in porous
media). Indeed, by considering creeping two-phase flow +\textbf{???} can
be written as

where \(p_i\) and \(u_i\) are microscopic pressure and velocity fields
of fluid \(i\), respectively. By expressing the surface integrals in
term of averaged quantities one can obtain the following momentum
balance equations,

in which \(\mathbf{K}_{ij}^{*}\) are the coupled relative permeability
tensors that pertain to the interaction between the fluids
{[}\protect\hyperlink{ref-Whitaker1986a}{10},\protect\hyperlink{ref-Lasseux1996}{11}{]}.

The reader is warned that +\textbf{???} has, to our knowledge, never
been used in practice to model two-phase subsurface flows. On the
contrary, the ubiquitous continuous model used to describe two-phase
flows in soils is based on a direct extension of the Darcy's equation
and the whole model, also known as Muskat equations
{[}\protect\hyperlink{ref-wyckoff1936flow}{12},\protect\hyperlink{ref-muskat1938flow}{13}{]},
reads

@@

where \(\mathbf{K}\) is the absolute permeability tensor. The
generalization toward two-phase flows involves the introduction of the
relative permeability terms \(k_{ri}\) which account for the division of
the void space between the fluids
{[}\protect\hyperlink{ref-dullien2012porous}{5}{]}, thus the relative
permeability depends (non-linearly) only on the saturation. To close the
set of the macroscopic equations a constitutive relation between the
macroscopic pressure of each fluid has to be furnished. This relation is
known as the capillary pressure relation and, as for the relative
permeabilities, is supposed to depends non-linearly only on the
saturation {[}\protect\hyperlink{ref-leverett1941capillary}{14}{]}.

A significant amount of work attempted to make improvements to the
Muskat equations (e.g.~including moving contact line
{[}\protect\hyperlink{ref-kalaydjian1992dynamic}{15}--\protect\hyperlink{ref-barenblatt2003mathematical}{17}{]}
or take into account the trapped phases
{[}\protect\hyperlink{ref-hilfer1998macroscopic}{18}{]}). Here we would
only stress that the concept of relative permeabilities into the Muskat
equations is dedicated to independent flow pathways only
{[}\protect\hyperlink{ref-blunt2017multiphase}{19}{]} and the underlying
assumption that the two streams do not interfere with each other, a
situation far from the regimes previously identified for high
permeability porous media. Thus +\textbf{???} does not take into account
the interaction between the fluids as opposed to +\textbf{???}. If the
former is used to the detriment of the latter, it is mainly because the
literature on subsurface flow is focused on oil recovery applications,
for which the permeability is very low and the independent streams
regime is a good approximation.

Another approach is to implement +\textbf{???} and provide constitutive
relations for the interaction terms between the phases. These relations
are usually obtained through interpretation of experimental data. Some
results, focusing on relation for the interaction between the fluid
phases, are given in the next section as well as results in literature
for coupled permeabilities from the porous medium approach.

\hypertarget{modeling-of-the-interaction-between-fluids}{%
\subsubsection{Modeling of the interaction between
fluids}\label{modeling-of-the-interaction-between-fluids}}

Based on the generalized Darcy law with coupled terms +\textbf{???},
experimental works computed the coupled transport coefficient through
steady-state cocurent flow in sandpack with one fluid, and alternatively
the other, which is submitted to a null pressure gradient. This protocol
was used with oil and water in a cylindric sandpack
{[}\protect\hyperlink{ref-zarcone1994determination}{20}{]} and the
authors found a negligible effect of the coupled permeabilities in the
overall flow. With the same protocol, with oil and water in a
2D-sandpack Dullien and Dong
{[}\protect\hyperlink{ref-Dullien1996}{21}{]} found that the coupled
permeabilities are important since they can contribute at best to 35\%
of the effective permeability. Alternatively, authors imposed
{[}\protect\hyperlink{ref-ramakrishnan2015measurement}{22}{]} a null
displacement and measured the induced pressure drop, with air and water
in a Berea sandstone core they found that the coupled transport
coefficients must not be overlooked for intermediate saturations.
Several authors conducted experiments with two different set-ups as
proposed by Rose {[}\protect\hyperlink{ref-Rose1988}{23}{]}, e.g.
{[}\protect\hyperlink{ref-bentsen1993use}{24}{]}, in which the authors
made cocurrent and countercurrent experiments with water and oil in a
sandpack and found that coupled permeabilities reach, at least, 15\% of
the effective permeability value. However, it was pointed out that the
saturation between the two sets of experiments can be very different and
therefore the computed relative permeabilities can not be safely
compared {[}\protect\hyperlink{ref-langaas2001numerical}{25}{]}. The
effect of the non-wetting phase connectivity on the transport parameters
was extensively studied in {[}\protect\hyperlink{ref-Avraam1995}{26}{]},
the authors performed steady-state cocurrent two-phase flow in 2D-micro
model experiments and found that the contribution of the coupled
permeabilities on the flow is non-negligible and depend on the flow
regimes.

Several numerical studies investigated the share of the coupled
permeabilities in the effective permeability. The seminal work of
Rothman {[}\protect\hyperlink{ref-Rothman1990}{27}{]} examined the
question by conducting two-phase flow simulations in simple geometries
with the immiscible lattice-gas method. The author found non-negligible
participation of the coupled permeabilities by applying the volume force
alternatively on each fluid. Numerous authors used the lattice-Boltzmann
method, such as {[}\protect\hyperlink{ref-Li2005}{28}{]}, in which the
authors examined the value of the coupled permeabilities as a function
of the saturation in a 3D-sphere pack and they found results in
agreement with Rothman's results. In
{[}\protect\hyperlink{ref-Yiotis2007}{29}{]} the authors used a lattice
Boltzmann method in 2D and 3D pore networks and found a non-wetting
apparent relative permeability greater than unity when the wetting fluid
is more viscous than the non-wetting fluid. This result is a
manifestation of the lubrification effect, also observed in experiments
{[}\protect\hyperlink{ref-odeh1959effect}{30}{]}, and arising because of
strong hydraulic coupling between the fluids. Recently, in
{[}\protect\hyperlink{ref-shams2018study}{31}{]}, the authors have used
a Volume Of Fluid method to study the transport coefficients of fluid
layers in non-circular capillary tubes. Based on analogy with a model
Couette flow the authors have derived simple relations that can predict
with good accuracy the transport coefficients, including the coupled
ones, of fluid layers. We can also refer to recent work in which a whole
analytical model is given for one-dimensional two-phase flow in coarse
non-consolidated porous media and which correctly predicts the pressure
loss in a debris bed
{[}\protect\hyperlink{ref-clavier2017modeling}{2}{]}.

Several correlations from theorical analysis for the fluid-fluid force
interaction have been derived in hydrodynamic of TBR litterature. These
correlations usually relate the momentum exchange between the fluids to
the flow velocity and the medium properties with Ergun's like relations.
For example, Attou et al. {[}\protect\hyperlink{ref-Attou1999}{32}{]}
found that the drag between the fluids can be expressed as follows

\[
F_{lg} = A_{lg} U_r + B_{lg} U_r^2,
\]

where the relative velocity is given by \(U_r=U_g-\frac{S_g}{S_l}U_l\)
and \(A_{lg}\) and \(B_{lg}\) are coefficients that depend on the volume
fraction of the gas phase and the medium properties. The subscripts
\(g\) and \(l\) denote the gas phase and the liquid phase, respectively.
This type of model with explicit account of interaction between fluids,
see also
{[}{[}\protect\hyperlink{ref-tung1988hydrodynamic}{33}{]};iliuta2005modelling{]},
gives good results when compared with experimental results
{[}\protect\hyperlink{ref-wang2013modelling}{34}{]}.

@@

@@

In the following, the superficial average of a physical quantity
\(\theta_i\) associated with the \(i\)-phase is given by

@@

Macroscopic quantities such as saturation and intrinsic average pressure
for two fluids \(i=w,o\) read

@@

As said, an essential step in order to derive some macroscopic law at
macroscale from the microscale is to use averaged quantities to express
the boundary intergals terms that arising because of the averaging
theorem application {[}\protect\hyperlink{ref-Whitaker2013}{35}{]}.
Remember that these closure relations are not presented here but rather
we leave intact the surface integrals, the averaged momentum balance
equations +\textbf{???} for two fluids \(w\) and \(o\), are recast as

@@

where \(\mathbf{T}_{ij}\) are the traction (name given to surface forces
in the Boundary Integral Methods literature in which these integrals are
central) exerted by fluid \(i\) upon fluid \(j\) and \(\mathbf{T}_{is}\)
the traction exerted by fluid \(i\) upon the solid
{[}\protect\hyperlink{ref-Kalaydjian1987}{9}{]}. The traction terms are
written as

@@

where \(\mathbf{n}_{ij}\) is the normal vector at the interface and
pointing toward the j-phase, \(p_{i}\) the pointwise pressure of the
fluid \(i\) and
\(\mathbf{e}_{i}=\frac{1}{2}\left(\nabla\mathbf{u}_{i}+\left(\nabla\mathbf{u}_{i}\right)^{T}\right)\)
the rate-of-strain tensor for Newtonian fluids with \(\mathbf{u}_{i}\)
the pointwise velocity of fluid \(i\).

@@

Two-phase flow between two parallel plates under study here is
illustrated in \cref{fig:modelStudy}. This represents a cocurrent flow
of two fluids inside a Hele-Shaw cell parallel to the \(x-y\) plane,
with the \(z\)-axis perpendicular to the plates, and for which \(h\) is
the aperture between the plates. A solid obstacle of circular
cross-section is sandwiched between the two plates. Tangential (cell
plane) stress, as well as stress arising from the perpendicular
confinement, need to be taking into account to properly describe the
flow. The important parameter here are the aspect ratio \(h/l\) where
\(l\) is some characteristic transverse length (hereafter \(l\) is the
cell width as shown in \cref{fig:modelStudy}) and the capillary number.

@@

To compute the traction exerted by the fluids upon the cell plates we
need to consider that the velocity profile in the \(z\)-direction is
well described by the Poiseuille flow profile,

@@

We note that the average velocity taken over the interval between the
plates is written as follows

@@

There is therefore a direct analogy between flows in a 2D porous medium
and flow in a Hele-Shaw cell, as can be seen in +\textbf{???}, where the
permeability is given by \(h^{2}/12\)
{[}\protect\hyperlink{ref-saffman1958penetration}{36}{]}.

where \(\gamma\) denotes the surface tension and \(\kappa_{\parallel}\)
the curvature in the Hele-Shaw plane
{[}\protect\hyperlink{ref-saffman1958penetration}{36}{]}. The
+\textbf{???} was completed in
{[}\protect\hyperlink{ref-park1984two}{37}{]} by an additional term that
pertains for dynamic film formation and a corrective \(\pi/4\) factor
for the curvature in the cell plane. The reader is warnes that while
dynamic film formation along the plates is an important phenomenom
regarding the flow {[}\protect\hyperlink{ref-jackson2015dynamic}{38}{]},
this requires dedicated study and is beyond the scope of this article.
Thus, we consider in the following that the invading fluid is in contact
with the cell plates, i.e.~presence of a triple line.

@@

In this equation, \(\mathbf{u}\) is the velocity field, \(\tau\) and
\(\psi\) are two numerical parameters that control the diffuse interface
thickness and the amount of reinitialization of the \(\phi\) function,
respectively
{[}\protect\hyperlink{ref-Olsson2005}{39},\protect\hyperlink{ref-Olsson2007}{40}{]}.

\hypertarget{flow-equations}{%
\subsubsection{Flow equations}\label{flow-equations}}

@@

where \(k=\frac{\sqrt{12}}{h}\) is an additional term that pertain to a
Darcy-term arising from the depth-averaging of the Stokes equation, as
described in the theoretical background section and \(\mathbf{F}_{st}\)
is the capillary contribution. The capillary contribution is difficult
to compute in volumetric methods such as Level Set and is itself the
object of a large body of work
{[}\protect\hyperlink{ref-Popinet2018}{41}{]}, this term is defined here
as

@@

where \(\gamma\) is the surface tension, \(\kappa\) is the interface
mean curvature, \(\mathbf{n}\) denotes the unit normal to the interface
and \(\delta(\phi)\) is the Dirac delta function localized on the
interface {[}\protect\hyperlink{ref-Galusinski2008}{42}{]}. In this
framework the unit normal vector as well as the mean curvature are
obtained with the level set function \(\phi\) as

@@

respectively, where the \(\pi/4\) correction was derived in
{[}\protect\hyperlink{ref-park1984two}{37}{]} and
\(\kappa_{\perp}=\frac{2}{h}\) is the out-of-plane curvature as
described previously in the theoretical background section.

@@

@@

\begin{longtable}[]{@{}lll@{}}
\caption{Equivalence between the value of the cell plates aperture and
the intrinsic permeability of the model obtained either from the
intrinsic permeability for flow between two plates or from single-phase
simulation. \label{tbl:permeability}}\tabularnewline
\toprule
\(l/h\) & \(h^2/12\) (darcy) & \(k\) Simulation (darcy)\tabularnewline
\midrule
\endfirsthead
\toprule
\(l/h\) & \(h^2/12\) (darcy) & \(k\) Simulation (darcy)\tabularnewline
\midrule
\endhead
0.2 & \(5.21 \times 10^{5}\) & \(1.48 \times 10^{4}\)\tabularnewline
2 & \(5.21 \times 10^{3}\) & \(2.85 \times 10^{3}\)\tabularnewline
4 & \(1.30 \times 10^{3}\) & \(8.63 \times 10^{2}\)\tabularnewline
8 & \(3.26 \times 10^{2}\) & \(2.35 \times 10^{2}\)\tabularnewline
20 & \(5.21 \times 10^{1}\) & \(3.94 \times 10^{1}\)\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{figure}
\hypertarget{fig:drag_plates}{%
\centering
\includegraphics{figures/pdf/dragPlates.pdf}
\caption{\(x\)-component of the traction on Hele-Shaw cell plates by
each fluid in UC3, as a function of the dimensionless gap and for
different capillary numbers at steady-state.}\label{fig:drag_plates}
}
\end{figure}
